---
title: "GT: Publication-Quality Tables from R"
subtitle:  <span style="color:#FFFFFF">twitter:@rinpharma github:philbowsher</span>
output:
  xaringan::moon_reader:
    css: "include/rstudio.css"
    nature:
      countIncrementalSlides: yes
      highlightLines: yes
      highlightStyle: github
      ratio: 16:9
resource_files:
  - include/rstudio.css
  - include/title.jpg
  - include/subtitle-white.png
  - include/panel-narrow.png
  - include/panel-wide.png
rsc_output_files:
  - include/rstudio.css
  - include/title.jpg
  - include/subtitle-white.png
  - include/panel-narrow.png
  - include/panel-wide.png
---


class: subtitle-slide, white, middle, center

# Thank you
[Richard Iannone](https://github.com/rich-iannone/),
[William Noble](https://www.linkedin.com/in/william-noble-8987a452/),
[Michael Stackhouse](https://www.linkedin.com/in/michael-s-stackhouse/),
[Sean Lopp](https://github.com/slopp),
[Shannon Hagerty](https://www.linkedin.com/in/shannonhagerty/)

---


class: panel-narrow-slide, left

# Slides and Examples

- [Github](https://github.com/philbowsher/phuse-GT-Package-for-Publication-Quality-Tables-from-R)
- [RStudio Cloud](https://rstudio.cloud/project/779965)

---



class: panel-narrow-slide, left

# Tables in Clinical Study Reports

- E3 Reporting
- Title and footnotes are integral parts of Tables, Figures or Listings (TFLs) in Clinical Study Reports (CSRs) and Case Report Tabulation (CRT)
- Titles and footnotes are created or typed in a requirement document such as Statistical Analyses Plan (SAP). 

---


class: panel-narrow-slide, left

# Main Topic of Interest for Pharma

- [rtf2](https://github.com/DanniYuGithub/RTF2/)
- Tables with complex headers
- RMD, HTML, [Interactivity](https://www.phusewiki.org/wiki/index.php?title=Best_Practices_for_Interactive_Analyses_for_Decision_Making_%26_Submissions)

---


class: panel-narrow-slide, left

# Managing Change
- Macros
- PROC REPORT, PROC TABULATE
- ODS
- [Roche diffdf](https://www.user2019.fr/static/pres/lt252425.pdf)

---



class: panel-narrow-slide, left

# But Why Another Table Package?

- [Listing of leading table-making R packages](https://gt.rstudio.com/)
- [How to Make Beautiful Tables in R](https://rfortherestofus.com/2019/11/how-to-make-beautiful-tables-in-r/)
- Different Pharma have unique formatting requirements
- General & flexible R package to customize table layout
- RStudio, Full time OS package developers maintainers

---




class: panel-narrow-slide, left


# Basic Example

```
devtools::install_github("rstudio/gt")
library(gt)

islands_tbl <-
  dplyr::tibble(
    name = names(islands),
    size = islands
  ) %>%
  dplyr::arrange(desc(size)) %>%
  dplyr::slice(1:10)
  
gt_tbl <- gt(data = islands_tbl) %>%  
  tab_header(
    title = "Large Landmasses of the World",
    subtitle = "The top ten largest are presented"
  ) %>%
  tab_source_note(
    source_note = "Source: The World Almanac and Book of Facts, 1975, page 406."
  ) %>%
  tab_source_note(
    source_note = md("Reference: McNeil, D. R. (1977) *Interactive Data Analysis*. Wiley.")
  )

gt_tbl
```

---







class: panel-narrow-slide, left

# GT
- Intro
- https://gt.rstudio.com/
- [RStudio Cloud](https://rstudio.cloud/project/779965)
- Demo

---










class: panel-narrow-slide, left

# Moderate Complexity OpenFDA Examples
- Adverse Events
- [Notebook](https://colorado.rstudio.com/rsc/content/6848/GT_Workflow.nb.html)
- [Dashobard](https://colorado.rstudio.com/rsc/content/6849/1-dashboard.html)
- Demo

---



class: panel-narrow-slide, left

```{r include=FALSE}

library(tidyverse)
library(openfda)
library(ggthemes)
library(gt)

get_adverse <- function(gender, brand_name) {
  fda_query("/drug/event.json") %>%
    fda_filter("patient.drug.openfda.brand_name", brand_name) %>% 
    fda_filter("patient.patientsex", gender) %>% 
    fda_count("patient.reaction.reactionmeddrapt.exact") %>% 
    fda_exec()
}

total_events<- function(brand_name){
  fda_query("/drug/event.json") %>%
    fda_filter("patient.drug.openfda.brand_name", brand_name) %>% 
    fda_count("patient.reaction.reactionmeddrapt.exact") %>% 
    fda_exec()
}

serious_events<- function(brand_name){
  fda_query("/drug/event.json") %>%
    fda_filter("patient.drug.openfda.brand_name", brand_name) %>% 
    fda_count("serious") %>% 
    fda_exec()
}

drug<- "Keytruda"
female<-get_adverse(2, drug) %>% mutate(sex = 'female')
male<-get_adverse(1, drug)%>% mutate(sex = 'male')

all<- total_events(drug) 
total_count<- sum(all$count)
serious<-serious_events(drug) %>% filter(term==1) %>% pull(count)

```

```{r echo = FALSE}
rbind(male,female) %>%  
  mutate(term=str_to_title(term)) %>% 
  pivot_wider(names_from=sex, values_from=count) %>% 
  mutate(Total=female+male) %>% 
  top_n(3, Total) %>% 
  gt(rowname_col="term") %>% tab_header(
    title = md("Adverse Events for Keytruda"),
    subtitle = "The top three most reported events"
) %>% tab_source_note("All data queried from openFDA") %>% 
  tab_stubhead(label= 'Patient Reaction') %>% 
  tab_spanner(label = "By sex",
              columns=vars(female, male)) %>% 
  tab_spanner(label= 'Total Cases',
              columns = vars(Total)) %>% 
    fmt_number(columns=vars(female, male, Total), sep_mark = ',', decimals=0) %>% 
  cols_label(Total = '',
             female = 'Female',
             male= 'Male')
```

---




class: panel-narrow-slide, left

# SDTM/ADaM Pilot Project
- [Atorus Research](https://atorusresearch.com/)
- [Pilot Submission Package](https://www.cdisc.org/sdtmadam-pilot-project)
- Xanomeline Transdermal
- Therapeutic System (TTS) in Patients with Mild to Moderate Alzheimer’s Disease

---







class: panel-narrow-slide, left

# Replicating 14-1.02 Summary of End of Study Data
- [htmltools](https://cran.r-project.org/web/packages/htmltools/index.html)
- [Table 14-1.02 Summary of End of Study Data](https://colorado.rstudio.com/rsc/content/6851/viewer-rpubs-718475a647c6.html)
- Demo
---












class: panel-narrow-slide, left



```{r include = FALSE}
library(dplyr)
library(glue)
library(tidyverse)
library(haven)
library(assertthat)

# library(pharmaRTF)
library(tibble)

source('~/CDISC_pilot_replication-master/programs/config.R')
source('~/CDISC_pilot_replication-master/programs/funcs.R')

#Read in Source and order factors
adsl <- read_xpt('~/CDISC_pilot_replication-master/data/adam/adsl.xpt')
adsl$COMP24FL <- ordered(adsl$COMP24FL, c("Y", "N", NA))
adsl$ARM <- ordered(adsl$ARM, c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"))
adsl$DCREASCD <- ordered(adsl$DCSREAS, c("Adverse Event",
                                         "Death",
                                         "Lack of Efficacy",
                                         "Lost to Follow-up",
                                         "Withdrew Consent",
                                         "Physician Decision",
                                         "I/E Not Met",
                                         "Protocol Violation",
                                         "Sponsor Decision"))


#### Completion Status Table
comp_stat <- adsl %>%
  group_by(COMP24FL, ARM) %>%
  summarise(n = n())

#Make data.frame for table, unnamed so the cols are named correctly
comp_df <- data.frame(
  "Placebo" = n_pct(unlist(comp_stat[c(1,4), "n"]), sum(unlist(comp_stat[c(1,4), "n"])), mark_lt=FALSE),
  "Xanomeline Low Dose" = n_pct(unlist(comp_stat[c(2,5), "n"]), sum(unlist(comp_stat[c(2,5), "n"])), mark_lt=FALSE),
  "Xanomeline High Dose" = n_pct(unlist(comp_stat[c(3,6), "n"]), sum(unlist(comp_stat[c(3,6), "n"])), mark_lt=FALSE),
  "Total" = c(n_pct(sum(comp_stat[1:3, "n"]), sum(comp_stat[,"n"]), mark_lt=FALSE),
              n_pct(sum(comp_stat[4:6, "n"]), sum(comp_stat[,"n"]), mark_lt=FALSE)),
  row.names = c("\tCompleted Week 24", "\tEarly Termination (prior to Week 24)"),
  #Stop data.frame from adding periods
  check.names = FALSE, stringsAsFactors = FALSE
)
# Add tabs to row.names

# Add missing row.
comp_df["\tMissing", ] <- "  0 (  0%)"

# p-value
comp_p <- fish_p(adsl, adsl$COMP24FL, adsl$ARM)
comp_df <- attach_p(comp_df, comp_p)


# sean version
# first, total by arm
arm_group <- adsl %>% 
  group_by(ARM) %>% 
  count()

# create table for early termination
term_df <- adsl %>% 
  filter(COMP24FL == "N") %>%
  group_by(DCREASCD, ARM) %>%
  mutate(DCREASCD = forcats::fct_expand(DCREASCD, "Missing")) %>% 
  count() %>% 
  ungroup() %>% 
  complete(DCREASCD, ARM, fill = list(n = 0)) %>% 
  rename(in_group = n) %>% 
  group_by(DCREASCD) %>% 
  add_tally(wt = in_group, name = "TOTAL") %>% 
  left_join(arm_group) %>% 
  mutate(pres_val = n_pct(in_group, n, mark_lt = FALSE)) %>%
  mutate(Total = n_pct(TOTAL, sum(n), mark_lt = FALSE)) %>% 
  select(DCREASCD, ARM, pres_val, Total) %>% 
  spread(ARM, pres_val) %>% 
  column_to_rownames("DCREASCD") %>% 
  relocate(Total, .after = last_col())

#### Reason for Early Termination Table
## By ARM
# term_reas <- adsl %>%
#   filter(COMP24FL == "N") %>%
#   group_by(DCREASCD, ARM) %>%
#   complete(nesting(DCREASCD, ARM)) %>%
#   #complete(DCREASCD, ARM) %>%
#   summarise(n = n())
# 
# ## Total
# term_reas_tot <- adsl %>%
#   filter(COMP24FL == "N", !is.na(DCDECOD)) %>%
#   group_by(DCREASCD) %>%
#   complete(nesting(DCREASCD, ARM)) %>%
#   summarise(n = n())
# 
# 
# term_df <- data.frame(
#   "Placebo" = n_pct(unlist(term_reas[seq(1, 27, 3), "n"]), sum(adsl %>% filter(ARM == "Placebo") %>% summarise(n = n()), na.rm=TRUE), mark_lt=FALSE),
#   "Xanomeline Low Dose" = n_pct(unlist(term_reas[seq(2, 27, 3), "n"]), sum(adsl %>% filter(ARM == "Xanomeline Low Dose") %>% summarise(n = n()), na.rm=TRUE), mark_lt=FALSE),
#   "Xanomeline High Dose" = n_pct(unlist(term_reas[seq(3, 27, 3), "n"]), sum(adsl %>% filter(ARM == "Xanomeline High Dose") %>% summarise(n = n()), na.rm=TRUE), mark_lt=FALSE),
#   "Total" = n_pct(unlist(term_reas_tot[, "n"]), sum(adsl %>% summarise(n = n())), mark_lt=FALSE),
#   row.names = c(
#     "\tAdverse Event",
#     "\tDeath",
#     "\tLack of Efficacy[2]",
#     "\tLost to Follow-up",
#     "\tSubject decided to withdraw",
#     "\tPhysician decided to withdraw subject",
#     "\tProtocol criteria not met",
#     "\tProtocol violation",
#     "\tSponsor decision"
#   ),
#   #Stop data.frame from adding periods
#   check.names = FALSE, stringsAsFactors = FALSE
# )
# term_df["\tMissing", ] <- "  0 (  0%)"



# p-value
term_p_1 <- adsl %>%
  select(ARM, DCREASCD) %>%
  mutate(loefl = ifelse(DCREASCD %in% "Adverse Event", 1, 0)) %>%
  fish_p(loefl, ARM, width = 6)
term_df <- attach_p(term_df, term_p_1)

term_p_2 <- adsl %>%
  select(ARM, DCREASCD) %>%
  mutate(loefl = ifelse(DCREASCD %in% "Lack of Efficacy", 1, 0)) %>%
  fish_p(ARM ,loefl, width = 6)
term_df$p[rownames(term_df) == "Lack of Efficacy"]  <- term_p_2



## Add Table labels
comp_df <- add_column(comp_df, " " = row.names(comp_df), .before = 1)
comp_df <- add_row(comp_df, " " = "Completion Status:", .before = 1)
comp_df <- add_row(comp_df, " " = "", .before = 1)

term_df <- add_column(term_df, " " = row.names(term_df), .before = 1)
term_df <- add_row(term_df, " " = "Reason for Early Termination (prior to Week 24):", .before = 1)
term_df <- add_row(term_df, " " = "", .before = 1)

combinedTable <- rbind(comp_df, term_df)
# Rename to get rid of period seperation
names(combinedTable)

headers <- adsl %>%
  group_by(ARM) %>%
  summarise(N = n())
headers_2 <- adsl %>%
  summarise(N = n()) %>%
  mutate(ARM = "Total")
headers_3 <- rbind(headers, headers_2) %>%
  mutate(labels = str_replace_all(str_wrap(glue('{ARM} (N={N})'), width=10), "\n", function(x) "\\line "))
headers_4 <- c(" ", headers_3$labels, "p-value [1]")
names(combinedTable) <- headers_4
```

```{r echo = FALSE}

library(gt)

colnames(combinedTable) <- colnames(combinedTable) %>% gsub("\\line ", "", ., fixed = TRUE)


four_line_header <- function(data, lines) {
  
  tab_header(
    data,
    title = html(
      htmltools::tagList(
        htmltools::tags$div(
          style = htmltools::css(padding.top = px(0), padding.bottom = px(0)),
          htmltools::tags$p(
            style = htmltools::css(
              text.align = "left", font.family = "Courier", font.size = px(14)),
            htmltools::tags$strong(lines[1]),
            htmltools::br(),
            htmltools::tags$strong(lines[2])
          ),
          htmltools::tags$p(
            style = htmltools::css(
              text.align = "center", font.family = "Courier", font.size = px(14)),
            htmltools::tags$strong(lines[3]),
            htmltools::br(),
            htmltools::tags$strong(lines[4])
          )
        )
      ) %>% as.character()
    )
  )
}




combinedTable %>%
  gt() %>%
  cols_label(
    `Placebo(N=86)` = html("Placebo<br>(N=86)"),
    `XanomelineLow Dose(N=84)` = html("XanomelineLow Dose<br>(N=84)"),
    `XanomelineHigh Dose(N=84)` = html("XanomelineHigh Dose<br>(N=84)"),
    `Total(N=254)` = html("Total<br>(N=254)")
  ) %>%
  four_line_header(
    lines = c("Protocol: CSIDSCPILOT01", "Population: Intent-to-Treat",
              "Table 14-1.02", "Summary of End of Study Data")
  ) %>%
  tab_style(
    locations = list(cells_body(), cells_column_labels(columns = TRUE)), 
    style = cell_text(font = "Courier", size = px(12))
  ) %>%
  tab_style(
    locations = cells_column_labels(columns = TRUE), 
    style = cell_text(weight = "bold")
  ) %>%
  fmt_missing(columns = everything(), missing_text = "") %>%
  opt_table_lines(extent = "none") %>%
  tab_options(
    column_labels.border.bottom.style = "solid",
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(1),
    table.width = px(1024), data_row.padding = px(2)
  ) %>%
  tab_source_note(
    source_note = html(
      htmltools::tagList(
        htmltools::tags$p(
          style = htmltools::css(font.family = "Courier", font.size = px(14)),
          "[1] Fisher's exact test.",
          htmltools::tags$br(),
          "[2] Based on either patient/caregiver perception or physician perception.",
        ),
        #htmltools::tags$br(),
        htmltools::tags$div(
          style = htmltools::css(font.family = "Courier", font.size = px(14)),
          htmltools::tags$span(
            style = htmltools::css(
              display = "block", float = "left", width = px(500)),
            "Source: C:\\csdic_pilot\\PROGRAMS\\DRAFT\\TFLs\\ads12.sas"
          ),
          htmltools::tags$span(
            style = htmltools::css(display = "block", float = "right", width = px(250)),
            "21:02 Monday, June 26, 2006"
          )
        )
      ) %>% as.character()
    ) 
  ) %>%
  tab_style(
    style = cell_text(indent = px(15)),
    locations = cells_body(columns = 1, rows = c(3:5, 8:17))
  )


```




---






















class: panel-narrow-slide, left

# RTF
- [pharmaRTF: Enhanced RTF Wrapper for Use with Existing Table Packages](https://github.com/atorus-research/pharmaRTF)
- [CDISC Pilot Replication in R](https://github.com/atorus-research/CDISC_pilot_replication)
- [r2rtf: Create RTF table and figure with flexible format.](https://merck.github.io/r2rtf/index.html) & [gsDesign](gsDesign: Group Sequential Design)
- [Typical Table Structure](https://merck.github.io/r2rtf/articles/example-pipeline.html)
- https://github.com/Merck/r2rtf/blob/master/vignettes/rtf/efficacy_example.rtf
- https://github.com/Merck/r2rtf/blob/master/vignettes/rtf/bs_pageby_example.rtf
---


class: panel-narrow-slide, left

# Shiny
- [The Future of Clinical TFLs is Interactive . . .](https://williamnoble.shinyapps.io/the_future_of_clinical_tfls/)
- [Genentech will be interested in conducting a real submission for the R Shiny application](https://www.phuse.eu/blog/working-groups-2019-end-of-year-summary)
- [scope is to develop tools to generate forest plots for inclusion in submissions to the FDA.](https://www.phuse.eu/standard-analyses-code-sharing)
- [FDA Shiny](https://www.cdisc.org/system/files/all/event/restricted/2019_Japan/16_AShapiro%2520CDISC%2520Japan%25202019%2520Presentation.pdf)
---

















class: panel-narrow-slide, left

# Collaboration

.pull-left[
  # [R in Pharma](https://rinpharma.com)
  ]
.pull-right[ 
  # [pharmar.org](https://pharmar.org)
]

---


class: panel-narrow-slide, left

# Questions?
phil@rstudio.com

